### 2011-11-06 doc/pldi2012/exec.tex

There are four kinds of names: functions $f$, regions $r$, local
variables $x$, and partitions $p$.  A function takes only regions as
arguments and a function body consists of an ordered list of
statements.  The core language has no permissions or coherence annotations
on functions; for simplicity, in this section we consider only exclusive access in the case
that a function writes a region.  The statement $f(r_1,\ldots,r_n)$ calls a function, which
is executed for its effects on the region arguments.  The statement $x
= \deref r$ reads a value from region $r$ and stores it in local
variable $x$, and the statement $\deref r = x$ stores the value of $x$
in region $r$.

@@

A program is a list of functions, with the entry point the first
function on the list.  We further require that every function (except
the entry point) is called in exactly one place and that the formal
parameters are named identically to the actual parameters in the
call.  Further, local variables, new region names and
partition names in {\tt partition} are chosen to be distinct from all
other names in the program.  These restrictions allow us to avoid
dealing with variable renaming and also mean that each runtime call is
uniquely named by its function name.

@@

The {\em sequential execution order} is the canonical order in which reads, writes, and partitions of regions are performed.
Th sequential order is obtained by inlining all function calls---repeatedly
replace function calls $g(r_1,\ldots,r_n)$ by their function bodies until the entry function contains no function calls.
The order of statements in the entry function is then the sequential execution order. 
Two programs $P_1$ and $P_2$ are equivalent, written $P_1 \equiv P_2$, if starting in the same initial states they halt in
the same final states when evaluated in the sequential execution order.

@@

\begin{lemma}
\rm
\label{lem:independence}
If $s_1$ and $s_2$ are independent, then $s_1; s_2 \equiv s_2; s_1$.
\end{lemma}
\begin{proof}
For a contradiction, assume that $s_1$ writes a region $r_1$ and $s_2$ may read from some $r_2$ where $r_1$ and $r_2$
are not disjoint; then reversing the order of the statements clearly violates the claim.  Now $r_1 \in \rwrite{s_1}$
and $r_2 \in \rread{s_2}$, and therefore $\lca{r_1}{r_2} = p$ for some partition $p$.  But then $r_1$ and $r_2$ are disjoint.
The reasoning is symmetric if $s_1$ reads $r_1$ and $s_2$ writes $r_2$ or $s_1$ writes $r_1$ and $s_2$ writes $r_2$.
Since the read-read case cannot violate independence, the result follows.
\end{proof}
We say $s_2$ {\em depends on} $s_1$ if $s_2$ follows $s_1$ in the sequential execution order and the statements are not independent.
\begin{lemma}
\rm
\label{lem:dataflow}
Let $S = s_1; \ldots s_n;$ be the sequential execution order of reads, writes, and partitions and let $S'$ be any permutation of $S$ such that the order of dependent statements is preserved.  Then $S \equiv S'$.
\end{lemma}
\begin{proof}
By induction on the number of swaps needed to transpose $S'$ into $S$.
If $S = S'$ then we are done.  If $S$ and $S'$ are different then
there must be two adjacent statements $s_1; s_2$ in $S'$ that are
reversed with respect to the sequential execution order.  Since the order of
dependent statements is preserved by the permutation, $s_1$ and $s_2$ are also independent.
But then $s_1; s_2 \equiv s_2; s_1$ by Lemma~\ref{lem:independence} and the result follows.
\end{proof}
Lemma~\ref{lem:dataflow} captures the standard idea that any
topological sort of the statements that preserves dependencies also
preserves sequential execution semantics.  While this is the basis for
dataflow parallelism, in the large distributed memory machines we
target instruction-level dataflow parallelism is both too fine-grain
and incurs too much communication, as we must compute a global
dependence graph across the entire program.  In Legion we use a
different notion of what can be executed in parallel that works at the
granularity of functions and also requires only local reasoning.  We
first need a few additional definitions.  Define $\context{s}$ to be
the function call statement that invokes the function in which
statement $s$ occurs (here we assume some way of distinguishing
identical statements that occur in different functions).  Let
$\interval{f}$ to be the set of statements executed by the one call to
function $f$ and all the functions $f$ transitively calls; we define
$\interval{s} = \{ s \}$ for any statement $s$ other than a function call.
Finally we say statements $s_1$ and $s_2$ are {\em siblings} if they
occur in the same function body.

\begin{lemma}
\rm
\label{lem:locality}
If $s_2$ and $s_1$ are not independent and $s_1$ and $s_2$ are not siblings, then $\context{s_2}$ and $s_1$ are not independent
and $s_2$ and $\context{s_1}$ are not independent.
\end{lemma}
\begin{proof}
Follows immediately from the fact that the read/writes sets of $\context{s}$ are supersetsn of the read/write sets of $s$.
\end{proof}
We use Lemma~\ref{lem:locality} in the following way.  Consider two
statements $s_1$ and $s_2$ such that $s_2$ depends on $s_1$.  Let $f$ be the function that is the least
common ancestor of $s_1$ and $s_2$ in the call tree of the
program---the unique function $f$ with the smallest interval such that
$s_1,s_2 \in \interval{f}$.  Then one of the following is true:
\begin{itemize}
\item $s_1$ and $s_2$ are siblings.

@@

\item There are two distinct function call statements $f_1(\ldots)$ and $f_2(\ldots)$ in the body of $f$ such
that $s_1 \in \interval{f_1}$ and $s_2 \in \interval{f_2}$ and $f_2(\ldots)$ is dependent on $f_1(\ldots)$.
\end{itemize}
In other words, any dependence between arbitrary statements
in different functions can be identified at a coarser granularity as a
dependence between two statements (one or both of which may be a function call) within the same function body.  For
example, in the example program above, statements 3 and 8 are
dependent because they write into regions that are in different
partitions of region $r_0$.  This dependence means that any functions
that include these statements in their interval cannot be independent,
and in particular in the entry point function $f$ the function call
$k(r_3,r_4)$ depends on sibling $g(r_1)$.

The following lemma gives another class of statement orderings based on these observations.
\begin{lemma}
\rm
\label{lem:scheduling}
Let $S$ be the sequential execution order of reads, writes and partitions.  Let $S'$ be any permutation of $S$ such that
for any sibling statements $s_1$ and $s_2$ such that $s_2$ depends on $s_1$, all statements in $\interval{s_1}$ precede
all statements in $\interval{s_2}$.  Then $S \equiv S'$.
\end{lemma}
\begin{proof}
It suffices to show that the order of any two dependent statements $s_1$ and $s_2$ is the same in $S$ and $S'$.
From the discussion above, we know that if $s_2$ depends on $s_1$ then there are two sibling statements $s_1'$ and $s_2'$
such that $s_1 \in \interval{s_1'}$, $s_2 \in \interval{s_2'}$, and $s_2'$ depends
on $s_1'$.
\end{proof}
The advantage of Lemma~\ref{lem:scheduling} is that provided we have
a way to efficiently obtain the read and write sets for function calls, scheduling
decisions can be made completely locally for the statements in a
single function body at a time, independently of any decisions for
other function calls.  This observation eliminates any global
computation to determine what statements can execute in parallel, and
furtherore allows decisions to be made in parallel, as the statements
in each function can be scheduled separately.  As a practical matter
our Legion implementation only parallelizes function calls---the other
statements within a function body are executed in sequential
order.

>>>>

There are four kinds of names: functions $f$, regions $r$, local
variables $x$, and partitions $p$.  A function takes only regions as
arguments and a function body consists of an ordered list of
statements.  The core language has no permissions or coherence annotations
on functions; for simplicity, in this section we simply use exclusive access in the case
that a function writes a region.  The statement $f(r_1,\ldots,r_n)$ calls a function, which
is executed for its effects on the region arguments.  The statement $x
= \deref r$ reads a value from region $r$ and stores it in local
variable $x$, and the statement $\deref r = x$ stores the value of $x$
in region $r$.

@@

A program is a list of functions, with the entry point the first
function on the list.  We require that every function (except
the entry point) is called in exactly one place and that the formal
parameters are named identically to the actual parameters in the
call.  Further, local variables, new region names and
partition names in {\tt partition} are chosen to be distinct from all
other names in the program.  These restrictions allow us to avoid
dealing with variable renaming and also mean that each runtime call is
uniquely named by its function name.

@@

The {\em sequential execution order} is the canonical order in which reads, writes, and partitions of regions are performed.
The sequential order is obtained by inlining all function calls---repeatedly
replace function calls $g(r_1,\ldots,r_n)$ by their function bodies until the entry function contains no function calls.
The order of statements in the entry function is then the sequential execution order. 
Two programs $P_1$ and $P_2$ are equivalent, written $P_1 \equiv P_2$, if starting in the same initial state they halt in
the same final state when evaluated in the sequential execution order.

@@

\begin{lemma}
\rm
\label{lem:independence}
If $s_1$ and $s_2$ are independent, then $s_1; s_2 \equiv s_2; s_1$.
\end{lemma}
\begin{proof}
For a contradiction, assume that $s_1$ writes a region $r_1$ and $s_2$ may read from some $r_2$ where $r_1$ and $r_2$
are not disjoint; then reversing the order of the statements clearly violates the claim.  Now $r_1 \in \rwrite{s_1}$
and $r_2 \in \rread{s_2}$, and therefore $\lca{r_1}{r_2} = p$ for some partition $p$.  But then $r_1$ and $r_2$ are disjoint.
The reasoning is symmetric if $s_1$ reads $r_1$ and $s_2$ writes $r_2$ or $s_1$ writes $r_1$ and $s_2$ writes $r_2$.
Since the read-read case cannot violate independence, the result follows.
\end{proof}
We say $s_2$ {\em depends on} $s_1$ if $s_2$ follows $s_1$ in the sequential execution order and the statements are not independent.
\begin{lemma}
\rm
\label{lem:dataflow}
Let $S = s_1; \ldots s_n;$ be the sequential execution order of reads, writes, and partitions and let $S'$ be any permutation of $S$ such that the order of dependent statements is preserved.  Then $S \equiv S'$.
\end{lemma}
\begin{proof}
By induction on the number of swaps needed to transpose $S'$ into $S$.
If $S = S'$ then we are done.  If $S$ and $S'$ are different then
there must be two adjacent statements $s_1; s_2$ in $S'$ that are
reversed with respect to the sequential execution order.  Since the order of
dependent statements is preserved by the permutation, $s_1$ and $s_2$ are also independent.
But then $s_1; s_2 \equiv s_2; s_1$ by Lemma~\ref{lem:independence} and the result follows.
\end{proof}
Lemma~\ref{lem:dataflow} captures the standard idea that any
topological sort of the statements that preserves dependencies also
preserves sequential execution semantics.  While this is the basis for
dataflow parallelism, in the large distributed memory machines we
target instruction-level dataflow parallelism is both too fine-grain
and incurs too much communication, as we must compute a global
dependence graph across the entire program.  In Legion we use a
different notion of what can be executed in parallel that works at the
granularity of functions and also requires analysis only within a function body.  We
first need a few additional definitions.  Define $\context{s}$ to be
the function call statement that invokes the function in which
statement $s$ occurs (here we assume some way of distinguishing
identical statements that occur in different functions).  Let
$\interval{f}$ be the set of statements executed by the one call to
function $f$ and all the functions $f$ transitively calls; we define
$\interval{s} = \{ s \}$ for any statement $s$ other than a function call.
Finally we say statements $s_1$ and $s_2$ are {\em siblings} if they
occur in the same function body.

\begin{lemma}
\rm
\label{lem:locality}
If $s_1$ and $s_2$ are not independent and not siblings, then $s_1$ and $\context{s_2}$ are not independent
and $s_2$ and $\context{s_1}$ are not independent.
\end{lemma}
\begin{proof}
Follows immediately from the fact that the read/writes sets of $\context{s}$ are supersets of the read/write sets of $s$.
\end{proof}
We use Lemma~\ref{lem:locality} in the following way.  Consider two
statements $s_1$ and $s_2$ such that $s_2$ depends on $s_1$.  Let $f$ be the function that is the least
common ancestor of $s_1$ and $s_2$ in the call tree of the
program---the unique function $f$ with the smallest interval such that
$s_1,s_2 \in \interval{f}$.  Then one of the following is true:
\begin{itemize}
\item $s_1$ and $s_2$ are siblings in $f$.

@@

\item There are two distinct function call statements $f_1(\ldots)$ and $f_2(\ldots)$ in the body of $f$ such
that $s_1 \in \interval{f_1}$ and $s_2 \in \interval{f_2}$ and $f_2(\ldots)$ is dependent on $f_1(\ldots)$.
\end{itemize}
In other words, any dependence between arbitrary statements
in different functions can be identified at a coarser granularity as a
dependence between two statements (one or both of which may be a function call) within the same function body.  
In the example program above, statements 3 and 8 are
dependent because they write into regions that are in different
partitions of region $r_0$.  This dependence means that any functions
that include these statements in their interval cannot be independent,
and in particular in the entry point function $f$ the function call
$k(r_3,r_4)$ depends on sibling $g(r_1)$.

The following lemma gives another class of statement orderings based on these observations.
\begin{lemma}
\rm
\label{lem:scheduling}
Let $S$ be the sequential execution order of reads, writes and partitions.  Let $S'$ be any permutation of $S$ such that
for any sibling statements $s_1$ and $s_2$ such that $s_2$ depends on $s_1$, all statements in $\interval{s_1}$ precede
all statements in $\interval{s_2}$.  Then $S \equiv S'$.
\end{lemma}
\begin{proof}
It suffices to show that the order of any two dependent statements $s_1$ and $s_2$ is the same in $S$ and $S'$.
From the discussion above, we know that if $s_2$ depends on $s_1$ then there are two sibling statements $s_1'$ and $s_2'$
such that $s_1 \in \interval{s_1'}$, $s_2 \in \interval{s_2'}$, and $s_2'$ depends
on $s_1'$.
\end{proof}
The advantage of Lemma~\ref{lem:scheduling} is that provided we have
a way to obtain the read and write sets for function calls, scheduling
decisions can be made completely locally for the statements in a
single function body at a time, independently of any decisions for
other functions.  This observation eliminates any global
computation to determine what statements can execute in parallel, and
furthermore allows scheduling decisions theselves to be parallelized, as the statements
in different functions can be scheduled separately.  As a practical matter
our Legion implementation only parallelizes function calls---the other
statements within a function body are executed in sequential
order.
